---
title: 'Historic redlining and biodiveristy observations'
author: "Amanda Herbst"
output: html_document
date: "2023-12-15"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE, message = FALSE, warning = FALSE)
```

## Background

Present-day environmental justice may reflect legacies of injustice in
the past. The United States has a long history of racial segregation
which is still visible. During the 1930s the Home Owners' Loan
Corporation (HOLC), as part of the New Deal, rated neighborhoods based
on their perceived safety for real estate investment. Their ranking
system, (A (green), B (blue), C (yellow), D (red)) was then used to
block access to loans for home ownership. Colloquially known as
"redlining", this practice has had widely-documented consequences not
only for community wealth, but also health.[^1] Redlined neighborhoods
have less greenery[^2] and are hotter than other neighborhoods.[^3]

[^1]: Gee, G. C. (2008). A multilevel analysis of the relationship
    between institutional and individual racial discrimination and
    health status. American journal of public health, 98(Supplement_1),
    S48-S56.

[^2]: Nardone, A., Rudolph, K. E., Morello-Frosch, R., & Casey, J. A.
    (2021). Redlines and greenspace: the relationship between historical
    redlining and 2010 greenspace across the United States.
    Environmental health perspectives, 129(1), 017006.

[^3]: Hoffman, J. S., Shandas, V., & Pendleton, N. (2020). The effects
    of historical housing policies on resident exposure to intra-urban
    heat: a study of 108 US urban areas. Climate, 8(1), 12.

Check out coverage by the [New York
Times](https://www.nytimes.com/interactive/2020/08/24/climate/racism-redlining-cities-global-warming.html).

A recent study found that redlining has not only affected the
environments communities are exposed to, it has also shaped our
observations of biodiversity.[^4] Community or citizen science, whereby
individuals share observations of species, is generating an enormous
volume of data. Ellis-Soto and co-authors found that redlined
neighborhoods remain the most undersampled areas across 195 US cities.
This gap is highly concerning, because conservation decisions are made
based on these data.

[^4]: Ellis-Soto, D., Chapman, M., & Locke, D. H. (2023). Historical
    redlining is associated with increasing geographical disparities in
    bird biodiversity sampling in the United States. Nature Human
    Behaviour, 1-9.

Check out coverage by
[EOS](https://eos.org/articles/bird-biodiversity-reports-reflect-cities-redlined-past?utm_source=EosBuzz&mkt_tok=OTg3LUlHVC01NzIAAAGOx3ckQPpeDY3QyFKdUnBSYqILj3qLrgb_Q2lasphwcF90xEzMgjZFUo3NHsw3caXDcZe6I8D8qdVsSRr3wzhb6LKAQRgNo8ffhjieEbDwMw).

## Workflow Goals

This workflow will investigate the lasting impacts of the HOLC grading system on the city of Los Angeles by:

- Using EJScreen data to map current environmental and demographic indicators in LA
- Comparing current conditions across census block groups in LA that were historically redlined
- Comparing percent of reported bird observations across areas in LA that were historically redlined

## Data Descriptions

### EJScreen

The United States Environmental Protection Agency's Environmental Justice Screening and
Mapping Tool is an effort to provide national data for advancing their goals and supporting others interested in environmental justice.

According to the [US EPA
website](https://www.epa.gov/ejscreen/purposes-and-uses-ejscreen):

> This screening tool and data may be of interest to community residents
> or other stakeholders as they search for environmental or demographic
> information. It can also support a wide range of research and policy
> goals. The public has used EJScreen in many different locations and in
> many different ways.
>
> EPA is sharing EJScreen with the public:\
> - to be more transparent about how we consider environmental justice
> in our work,\
> - to assist our stakeholders in making informed decisions about
> pursuing environmental justice and,\
> - to create a common starting point between the agency and the public
> when looking at issues related to environmental justice.

EJScreen provides environmental and demographic information for the
US at the Census [tract](https://en.wikipedia.org/wiki/Census_tract) and
[block group](https://en.wikipedia.org/wiki/Census_block_group) levels. Block group data has been downloaded, and stored locally, from
the [EPA site](https://www.epa.gov/ejscreen/download-ejscreen-data).

### Mapping Inequality

A team of researchers, led by the [Digital Scholarship
Lab](https://dsl.richmond.edu/) at the University of Richmond have
digitized maps and information from the HOLC as part of the [Mapping
Inequality](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58)
project.

This workflow uses maps of HOLC grade designations for Los Angeles via URL.
Information on the data can be found
[here](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58&text=downloads).[^5]

[^5]: Robert K. Nelson, LaDale Winling, Richard Marciano, Nathan
    Connolly, et al., "Mapping Inequality," American Panorama, ed.
    Robert K. Nelson and Edward L. Ayers, accessed October 17, 2023,
    <https://dsl.richmond.edu/panorama/redlining/>

### Biodiversity observations

The [Global Biodiversity Information Facility](gbif.org) is the largest
aggregator of biodiversity observations in the world. Observations
typically include a location and date that a species was observed. This workflow will use bird observations in 2022.

## Legacy of Redlining Workflow

### 1. Investigate the legacy of redlining in current environmental (in)justice

#### Import libraries and data

```{r }
library(tidyverse)
library(sf)
library(plotly)
library(gt)
```

Since we are analyzing LA in this workflow, we will read in EJScreen data and filter to Los Angeles County in the same step so we have a much smaller dataset to start with.

```{r }
# read in ejscreen data with direct filepath
la_county <- st_read("~/Documents/github/eds223-assignments/assignment-2-amandaherbst/data/EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb/",
                    quiet = TRUE) %>% 
  # filter ejscreen to LA county
  filter(CNTY_NAME == "Los Angeles County")
```

#### Understanding present day environmental justice issues in LA

**Wastewater Discharge**

The wastewater discharge indicator in EJScreen assigns a score to each census block group based on the pollution and toxicity of nearby streams and rivers. LA has limited access to local clean water and often deals with extremely polluted water after a storm. Let's visualize the wastewater discharge indicator for LA census block groups and add centroids for block groups that are above the 95th percentile of national values for wastewater discharge. (If this is taking too long, don't run the `ggplotly` line, it is not necessary, just makes the map interactive.) If you want to save the output, uncomment the last line!

```{r }
# map of LA
# color census block groups by wastewater discharge
# add centroid for block groups above 95th percentile

# filter la dataset for block groups above 95th percentile for wastewater dischare
top_percentile <- la_county %>% 
  filter(P_PWDIS > 95)

# make centroids for block groups above 95th percentile
top_percentile_centroids <- st_centroid(top_percentile)

# map LA county and color census blocks by waste water discharge
m1 <- ggplot() +
  geom_sf(data = la_county, aes(fill = PWDIS),
          lwd = 0.1,
          color = 'gray') +
  scale_fill_viridis_c(direction = -1) +
  # add above 95th percentile centroids
  geom_sf(data = top_percentile_centroids,
          color = 'magenta') +
  labs(x = "Longtitude",
       y = "Latitude",
       fill = "Wastewater discharge") +
  # choose x axis breaks so longitude is readable
  scale_x_continuous(breaks = c(-119, -118.6, -118.2, -117.8)) +
  theme_bw()

# make interactive to better see where the centroids are
ggplotly(m1)

# save plot in Outputs folder
#ggsave("la_wastewater_discharge.png", plot = m1, path = "../Outputs")
```



**In numbers**

Now let's use EJScreen data to investigate demographic (% low income) and environmental (PM 2.5 and Superfund proximity) indicators. 

We can calculate what percent of census block groups have less than 5% of the
population considered low income by filtering to groups with a percent income less than 0.05. Each row corresponds to a block group so the percent can be calculated by dividing the number of rows of the filtered dataframe by the number of rows of the original dataframe * 100.

```{r }
high_inc <- la_county %>% 
  # filter to block groups where % of low income is less than 0.05
  filter(LOWINCPCT < 0.05) %>% 
  # take the resulting number of rows and divide by the total number of block groups
  # multiply by 100 to get percent value
  nrow()/nrow(la_county) * 100

print(paste(round(high_inc,2), "% of LA census block groups have less than 5% of the population considered low income"))
```

We can calculate the percent of census block groups that are above the 80th
percentile for Particulate Matter 2.5 AND above the 80th percentile for
Superfund proximity in a similar way as above. First, the ejscreen data is filtered where the percentiles for PM 2.5 and Superfund proximity are both over 80, then those number of rows are divided by the number of rows of the original dataframe * 100.

```{r }
# percent of census block groups that are above 80th percentile for both PM 2.5 and Superfund proximity
pm_sf <- la_county %>% 
  filter(P_PM25 > 80 & P_PNPL > 80) %>% 
  nrow() / nrow(la_county) * 100

print(paste(round(pm_sf,2), "% of LA census block groups are above the 80th percentile for both PM 2.5 and Superfund proximity"))
```

#### LA redlining

Maps of HOLC grade designations for Los Angeles can be imported from the Mapping Inequality project via URL as an `sf` object and geometries are made valid with `st_make_valid`.

```{r }
LA_redlining <- st_read("https://dsl.richmond.edu/panorama/redlining/static/citiesData/CALosAngeles1939/geojson.json",
                        quiet = TRUE) %>%
  st_make_valid()
```

We can explore the redlining data by plotting the historical boundaries in LA, indicating HOLC grade
by color. If you want to save the output, uncomment the last line!

```{r }
# plot redlining areas, color by HOLC Grade
m2 <- ggplot() +
  geom_sf(data = LA_redlining, aes(fill = grade)) +
  labs(x = "Longtitude",
       y = "Latitude",
       fill = "HOLC Grade") +
  theme_bw()

m2

# save map in Outputs folder
#ggsave("la_holc_census_block_grps.png", plot = m2, path = "../Outputs")
```

To investigate current conditions in historically redlined areas, we can find the number of census block groups that fall within areas with HOLC grades. First, the `LA_redlining` and `la_county` data are in different coordinate reference systems, so one is reprojected before continuing. Then, a spatial inner join can be performed between the two datasets so that the resulting dataset contains only census block groups that are *within* areas with a historic HOLC grade.

```{r}
# check CRS!
st_crs(la_county) == st_crs(LA_redlining)
# la_county is in EPSG:3857
# LA_redlinging is in EPSG:4326
```

```{r}
# transfrom LA_redlining to same crs as la_county
LA_redlining3857 <- st_transform(LA_redlining, st_crs(la_county))

# block groups that have intersecting geometries as LA_redlining HOLC grades
# left = FALSE for an inner join
block_groups_holc_join <- st_join(x = la_county, y = LA_redlining3857, left = FALSE)
holc_block_grps <- nrow(block_groups_holc_join)

print(paste(holc_block_grps, "census block groups in LA fall within areas with HOLC grades"))
```

Based on EJScreen data, we can now summarize current conditions in the census block
groups that were within areas with a HOLC grade. We will group by HOLC
grade to compare the conditions across areas that were historically
graded A, B, C, and D. Let's look at the average % low income,
percentile for particulate matter 2.5, percentile for low life
expectancy, and percentile for air toxics cancer risk. If you want to save the output, uncomment the last line!

```{r }
# summary table of means grouped by grade
t1 <- block_groups_holc_join %>%
  st_drop_geometry() %>% 
  group_by(grade) %>% 
  summarise(low_inc_pct_avg = mean(LOWINCPCT, na.rm = TRUE),
            p_pm25_avg = mean(P_PM25, na.rm = TRUE),
            p_low_life_exp_avg = mean(P_LIFEEXPPCT, na.rm = TRUE),
            p_cancer_avg = mean(P_CANCER, na.rm = TRUE)) %>% 
  gt() %>% 
  tab_header(title = "Current EJ Conditions in LA by HOLC grade")

t1

# save table as image in Outputs folder
#gtsave(t1, "la_current_conditions_tbl.png", path = "../Outputs")
```

Based on the averages calculated, the HOLC grade for a census block
group in LA appears to be indicative of the percent of the population
that is low income, as well as the percentiles they fall within for
environmental justice indicators. Groups with a grade of "A" had the
lowest average percent of low income (15%), and had an average of the
lowest percentiles, comparatively, for PM 2.5, low life expectancy, and
air toxics cancer risk. In comparison, areas with a grade of "D" had the
highest average percent of low income (39%) and had the highest average
percentiles for PM 2.5, low life expectancy, and air toxics cancer risk.

### Investigate the legacy of redlining in biodiversity observations

To add bird biodiveristy observations to the analysis, import citizen science bird observations from a direct filepath and select observations from the year 2022. With another spatial inner join between
`LA_redlining` and `la_birds_22` (after checking for matching CRS's), filter for observations that fall
within neighborhoods with HOLC grades. 

```{r }
# read in birds data
la_birds <- st_read("~/Documents/github/eds223-assignments/assignment-2-amandaherbst/data/gbif-birds-LA",
                    quiet = TRUE)

# filter birds to 2022
la_birds_22 <- la_birds %>% 
  filter(year == 2022)
```

```{r}
# Check crs!
st_crs(la_birds_22) == st_crs(LA_redlining)
# both EPSG:4326
```

Matching CRS's means we can continue without having to reproject!

Once joined, the percent of bird observations within each redlining category can be calculated by grouping by HOLC grade, then dividing the number of observations in each grade by the total number of observations (number of rows in original joined dataframe). Visualize these percentages with a bar chart for readability. If you want to save the output, uncomment the last line!

```{r}
# bird observations in neighborhoods with HOLC grades
# inner join using "left = FALSE"
holc_birds <- st_join(LA_redlining, la_birds_22, left = FALSE)

# total number of bird observations
tot_bird <- nrow(holc_birds)

# percent of bird observations by HOLC grade
holc_birds_pct <- holc_birds %>% 
  group_by(grade) %>% 
  summarize(bird_pct_obs = n()/tot_bird * 100) 


# plot % of bird observations by HOLC
p1 <- ggplot(data = holc_birds_pct) +
  geom_col(aes(x = grade, y = bird_pct_obs, fill = grade),
           show.legend = FALSE) +
  labs(x = "HOLC Grade",
       y = "% Bird Observations",
       title = "Bird Observations in Redlining Categories") +
  theme_bw()

p1 

#ggsave("bird_obs_by_holc_grade.png", plot = p1, path = "../Outputs")
```

Since Ellis-Soto and co-authors found that redlined neighborhoods remain
the most undersampled areas across 195 US cities, it is surprising that
the largest share of bird observations are in historically redlined
areas of Los Angeles. This could potentially be a result of LA's
population skyrocketing since 1933 and consequently leading to
gentrification of redlined neighborhoods.
